---
title: "Lab 5"
author: "Anders Lie"
date: "4/16/2021"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(rmarkdown)
library(lubridate)
library(ggrepel)
states <- map_data("state")
state_names <- states %>% group_by(region) %>% summarize(long=mean(long), lat=mean(lat))
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
ppl <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
glc <- readxl::read_xlsx("FRPP_GLC.xlsx")
```

# Preprocessing
```{r}
# Preprocess GLC
glc$STATE = as.numeric(glc$`State Code`)

glc_states <- glc %>% group_by(STATE) %>% summarize(state_name=first(`State Name`))
acc_with_state_names <- acc %>% left_join(glc_states, by=c('STATE'))
```

# 1
```{r}
acc %>% group_by(DAY_WEEK) %>%
  summarize(number_accidents=n()) %>%
  mutate(weekday=c('Sun', 'Mon', 'Tue', 'Wed', 'Thu', 'Fri', 'Sat')) %>%
  select(-DAY_WEEK) %>%
  subset(select=c(2,1))
```

Using the data manual, we know that values of DAY_WEEK 1-7 corresponds to days, Sun-Sat, which we
can use to get the weekday label for each integer representation. In the resulting table,
we can see that on Fridays, Saturdays, and Sundays, there are generally a significantly 
greater number of fatal accidents compared to Monday, Tuesday, Wednesday, or Thursday.

The greatest number of fatal accidents occur on Saturdays.


# 2
```{r}
fatal <- ppl %>% filter(INJ_SEV==4)
head(fatal)
```

Here we see a few entries of the dataset created which contains only the people who suffered fatal injuries.

# 3
```{r}
# First get the MAKE for each accident
dat <- acc_with_state_names %>% inner_join(ppl, by=c('ST_CASE', 'STATE')) %>% select(STATE, state_name, MAKE)

most_dangerous <- dat %>%
  filter(!is.na(MAKE)) %>%
  group_by(STATE, state_name, MAKE) %>%
  summarize(number_accidents=n()) %>%
  filter(number_accidents==max(number_accidents))
most_dangerous %>% paged_table
```
Here we can see the most dangerous make ID for each state ID, as well as the number of 
fatal accidents for each of those state/make combinations.

# 4
```{r}
states$region = toupper(states$region)

avg_state_locations <- states %>% group_by(region) %>% summarize(avg_long=mean(long), avg_lat=mean(lat))

dangerous_with_location <- most_dangerous %>% mutate(region=state_name) %>% inner_join(avg_state_locations, by=c('region'))

ggplot(states, aes(x=long, y=lat)) + geom_polygon(aes(group=group)) +
  geom_text_repel(data=dangerous_with_location, aes(x=avg_long, y=avg_lat, label=MAKE), color='green')
```

Here we can see the most dangerous vehicle (most fatal accidents) make code for each state.

# 5
```{r}
acc_ppl <- acc %>% inner_join(ppl)
paged_table(acc_ppl)
```

Show above are a few entries of the joined dataframe. ST_CASE identifies each accident, so we could join by that variable,
but we can also just allow R to match by all matching columns to get an equivalent result.

# 6
```{r}
acc %>% group_by(DAY_WEEK) %>%
  summarize(number_accidents=n()) %>%
  mutate(weekday=c('Sun', 'Mon', 'Tue', 'Wed', 'Thu', 'Fri', 'Sat')) %>%
  select(-DAY_WEEK) %>%
  subset(select=c(2,1)) %>%
  ggplot(aes(x=weekday, weight=number_accidents)) + geom_bar() +
  xlab('Day of Week') + ylab('Number of Accidents')
  

acc %>% group_by(HOUR) %>%
  filter(HOUR != 99) %>%
  summarize(number_accidents=n()) %>%
  ggplot(aes(x=HOUR, weight=number_accidents)) + geom_bar() +
  xlab('Hour in Day') + ylab('Number of Accidents')

acc_ppl %>% group_by(SEX) %>%
  summarize(number_accidents=n()) %>%
  ggplot(aes(x=SEX, weight=number_accidents)) + geom_bar() +
  xlab('Sex Attribute Code') + ylab('Number of Accidents')
```

We can see from these summaries that Saturday is the most common day on which fatal accidents occur,
fatal accidents are most common around hour 18 (6 PM), and that males (Sex Attribute 1) are about twice
as likely to be in an accident that females (Sex Attribute 2).

# 7
```{r}
counties <- map_data('county')

deaths_by_county <- acc_ppl %>% 
  filter(INJ_SEV==4) %>% 
  group_by(STATE, COUNTY) %>%
  summarize(total_deaths=n())

glc_counties <- glc %>%
  mutate(STATE=as.numeric(`State Code`), COUNTY=as.numeric(`County Code`), region=tolower(`State Name`), subregion=tolower(`County Name`)) %>%
  group_by(region, subregion, STATE, COUNTY) %>%
  summarize(region=first(region), subregion=first(subregion), STATE=first(STATE), COUNTY=first(COUNTY))

#glc_loc %>% glc_counties %>% left_join()
#glc_counties
#deaths_by_county
#counties

deaths_by_county_with_names <- deaths_by_county %>% left_join(glc_counties, by=c('STATE', 'COUNTY'))
deaths_by_county_with_names

#counties

dat <- counties %>% left_join(deaths_by_county_with_names, by=c('region', 'subregion'))
#dat <- deaths_by_county_with_names %>% left_join(counties, by=c('region', 'subregion'))

head(deaths_by_county_with_names)
head(counties)
head(dat)


#counties_with_codes <- counties %>% inner_join(glc_counties, by=c('region', 'subregion'))
#counties_with_codes
#dat
ggplot(dat, aes(x=long, y=lat, fill=total_deaths)) + geom_polygon(aes(group=group))
```

Above we can see the total deaths due to accidents by county. Grayed out counties had
no records in the accident dataset, suggesting that most likely they have had no recorded
fatal accidents.

# 8
```{r}
summer <- acc_with_state_names %>% filter(MONTH %in% c(6,7,8))
winter <- acc_with_state_names %>% filter(MONTH %in% c(12, 1, 2))

nrow(summer)
nrow(winter)
```

We can see that there are more fatal accidents during the summer, suggesting that
the summer is more dangerous than the winter in general.

```{r}
summer1 <- summer %>% group_by(state_name) %>% summarize(num_summer_accidents=n())
winter1 <- winter %>% group_by(state_name) %>% summarize(num_winter_accidents=n())

dat <- inner_join(summer1, winter1)
dat %>% filter(num_winter_accidents > num_summer_accidents) %>% paged_table
```

We can see that this does, however, depend on the state. The table above shows
exactly which states for which there are more fatal accidents in the winter
than in the summer.