---
title: "Lab 5"
author: "Anders Lie"
date: "4/16/2021"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(rmarkdown)
library(lubridate)
states <- map_data("state")
state_names <- states %>% group_by(region) %>% summarize(long=mean(long), lat=mean(lat))
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
ppl <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
glc <- readxl::read_xlsx("FRPP_GLC.xlsx")
```

# 1
```{r}
acc %>% group_by(DAY_WEEK) %>%
  summarize(number_accidents=n()) %>%
  mutate(weekday=c('Sun', 'Mon', 'Tue', 'Wed', 'Thu', 'Fri', 'Sat')) %>%
  select(-DAY_WEEK) %>%
  subset(select=c(2,1))
```

Using the data manual, we know that values of DAY_WEEK 1-7 corresponds to days, Sun-Sat, which we
can use to get the weekday label for each integer representation. In the resulting table,
we can see that on Fridays, Saturdays, and Sundays, there are generally a significantly 
greater number of fatal accidents compared to Monday, Tuesday, Wednesday, or Thursday.

The greatest number of fatal accidents occur on Saturdays.


# 2
```{r}
fatal <- ppl %>% filter(INJ_SEV==4)
head(fatal)
```

Here we see a few entries of the dataset created which contains only the people who suffered fatal injuries.

# 3
```{r}
# First get the MAKE for each accident
dat <- acc %>% inner_join(ppl, by=c('ST_CASE', 'STATE')) %>% select(STATE, MAKE)
dat

dat %>%
  filter(!is.na(MAKE)) %>%
  group_by(STATE, MAKE) %>%
  summarize(number_accidents=n()) %>%
  filter(number_accidents==max(number_accidents))
```
Here we can see the most dangerous make ID for each state ID, as well as the number of 
fatal accidents for each of those state/make combinations.

# 4
```{r}
ggplot(states, aes(x=long, y=lat)) + geom_polygon(aes(group=group))
```
